| Early stopping in clinical PET studies: \n| how to reduce expense and exposure

### 2020-10-16 ArticleDraft/Draft.Rmd
Last fixes for version 2 on  medRxiv

Traditionally, the number of included subjects in a clinical PET study is determined a priori by a power analysis based on the null-hypothesis-significance testing (NHST) procedure[^fn2] [@Perezgonzalez2015]. Following this, subjects are recruited and scanned until the specified sample size is reached. In the case of a patient-control comparison study, a p-value for the mean difference of an outcome in a region of interest is then calculated. This value reflects the plausibility of obtaining the observed mean difference, or a more extreme difference, given that the null hypothesis (H~0~) is true (i.e., no difference between the groups). If the p-value is above a predefined alpha threshold (usually set to 0.05), the result is interpreted as inconclusive and H~0~ is not rejected (nor may it be confirmed!). The researcher might have committed a type II error, in which case more data would likely be needed to detect a potential difference. Alternatively, H~0~ might be true, i.e., there is no population difference to detect. If the p-value is below the alpha threshold, the result is instead interpreted as there being a significant difference between groups, inferring that there is an actual difference at the population level. Although many PET-studies likely suffer from insufficient power because of limited sample sizes, it is also important to note that a difference between groups may often be detectable before reaching the sample size that was determined a priori. Ideally, the researcher should include just the number of individuals needed to reach a correct conclusion, no more, no less. When performing more scans than needed, the PET researchers are wasting money and exposing people unnecessarily to radioactivity.

@@

If a researcher continues to collect data indefinitely, the default BF t-test will eventually converge to supporting the hypothesis most compatible with the population effect, i.e., it will show evidence for the null-hypothesis if in fact there is no mean difference, or for the alternative-hypothesis if there is an underlying true difference in the population [@Rouder2009]. However, such continuous collection of data is often unrealistic in a clinical PET study. Apart from the high cost, a PET examination also entails injecting subjects with radioactivity. Hence, a maximum sample size usually has to be decided on a priori and approved by an ethical review board and/or a radiation safety committee.

@@

Figure \ref{fig_k_at_5fpr}A and 3B show the estimated false positive rate when applying sequential BF testing in the case in which there is no difference between the two groups. Three different stopping thresholds have been used (BF > 3, 4 and 6, respectively). When the maximum allowed sample size increases, the false positive rate goes up. This is because the longer the sequential testing can go on, the more decisions are being made. Some of these decisions will be wrong, meaning that the researcher will claim support for an effect, when in fact there is no effect in the population. Eventuelly, when N becomes high enough, the rate of false positives will reach an asymptote (see Supplementary Figure S1). But with a maximum sample size above 30 subjects/group, it is not possible to keep false positives below the 5% using a decision threshold of 4. Instead, if the researchers wish to control the false positives at the 5% level, one option is to increase the decision threshold. Figure \ref{fig_k_at_5fpr}C and 3D show the estimated decision threshold that would be needed to keep the false positive rate at 5% when using sequential testing with the default BF t-test, while still starting at 12 subjects/group and checking the BF after the collection of each patient-control pair.

@@

##Objective 2 - stopping for both H~1~ and H~0~
The second aim of this tutorial is to assess whether sequential BF testing can be used to stop a clinical PET study early, both when there is an effect in the population, but also when the population effect is zero. After the collection of each data point, we will compare both the BF, and its reciprocal, to an a priori set decision threshold. If either the BF or its reciprocal passes the threshold, the study will be stopped, and we will declare support in favour of the alternative or null hypothesis, respectively.

###Simulation set-up 
We used the same simulation set-up as above, where a grid of maximum sample sizes and effect sizes ranging from 0 to 1.5 was used to evaluate sequential BF testing in a cross-sectional and paired design respectively. We used the same settings for the default BF t-test as above with one exception: the alternative hypothesis was still a Cauchy distribution (centered at zero with r=0.707) but now specified as being one-sided, instead of two-sided. This means that we anticipate that the effect will go in one direction (e.g., patients will have a higher BP~ND~ than controls), making the test into a one-sided default BF t-test. The reason for considering only a one-sided test in this scenario is that it is not possible to stop for H~0~ when using a two-sided test at smaller sample sizes when using reasonable decision thresholds (BF < $\nicefrac{1}{4}$, see supplementary Figure S11).

@@

\begin{tcolorbox}[title=\textbf{Practical recommendations 2},title filled]
{In this simulation set-up we show that it is possible to stop a PET study early either when there is a true underlying effect in the population, or when the effect is zero. For a patient-control comparison, we recommend that sequential BF testing not start until at least 12 subjects per group have been collected, setting the BF threshold to a minimum of 3, and using a one-sided Cauchy distribution (centered around zero with a \textit{r} = 0.707) as the alternative hypothesis. Using a one-sided test means that the researcher must make an a priori prediction of which direction the effect will have, and not change this prediction after the study is started, similar to performing a classical one-sided NHST t-test.}
\end{tcolorbox}

@@

#### Evidence in favor of H~0~
One advantage of sequential BF testing is that a PET study can also be stopped early when there is no effect in the population, i.e., when H~0~ is true. If the researcher wants to use a sequential BF to assess support for both H~1~ and H~0~, we recommend using a one-tailed default BF t-test. This entails that the researcher must make an a priori prediction on the direction of the effect (e.g., patients have higher binding than control subjects, or vice versa). Without using one-sided default BF t-test, it will be difficult (or even impossible) to stop a study early in support for H~0~, given the sample sizes used in typical PET studies (N<100 subjects/group) while still using a threshold that is high enough to keep the rate of false evidence at reasonable levels (see Supplementary Figure S12).

@@

1. Increase the a priori set decision threshold for stopping (see Figure \ref{fig_k_at_5fpr}).
2. Recruit and scan a larger pool of subjects (>12/group) before initiating the sequential BF testing. 
3. Check BF less frequently, e.g., only after the collection of each second or third (etc.) comparison pair, instead of after each single one.

@@

Figure \ref{fig_powerPhase2} presents the true and false evidence rates the researchers can expect when using a one-sided BF t-test and when they are correct in their prediction of the direction of an effect. If there is a sizable patient-control difference in the population, but it goes in the opposite direction to the predicted one, then using this approach would lead studies to be stopped early due to the null-hypothesis being supported almost 100% of the time. Of course, the null is not true in this case, but it is closer to the true population effect than the a priori defined alternative hypothesis. Hence, it will often gain the most support from the data.

@@

#### Inconclusive findings 
For any given PET-study where sequential BF testing is applied there will be a chance of reaching Nmax without crossing the pre-set decision threshold. In such cases, where the researcher ends up with an “inconclusive” finding, the BF is still interpretable. The suggested evidence thresholds in Table 1 can be used to report the support in data in favor of the two competing hypotheses, even though BF never formally reached the a priori set stopping threshold. For example, if the BF is above 3 (or below $\nicefrac{1}{3}$), this can still be reported as “moderate” support in data for one hypothesis over the other. However, in doing so the researcher cannot any longer say that they are controlling the rate of false evidence at a prespecified level.

@@

If a PET researcher wants to perform a study using sequential BF testning, we recommend that they follow the steps outlined in a flow-chart found in supplementary information (supplementary Figure S12). In order to perform the default BF t-test, the freely available *BayesFactor* package [@Morey2018] in R or point-and-click software JASP [@Quintana2018; @Love2019; @VanDoorn2019] can be used.

>>>>

Traditionally, the number of included subjects in a clinical PET study is determined *a priori* by a power analysis based on the null-hypothesis-significance testing (NHST) procedure[^fn2] [@Perezgonzalez2015]. Following this, subjects are recruited and scanned until the specified sample size is reached. In the case of a patient-control comparison study, a p-value for the mean difference of an outcome in a region of interest is then calculated. This value reflects the plausibility of obtaining the observed mean difference, or a more extreme difference, given that the null hypothesis (H~0~) is true (i.e., no difference between the groups). If the p-value is above a predefined alpha threshold (usually set to 0.05), the result is interpreted as inconclusive and H~0~ is not rejected (nor may it be confirmed!). The researcher might have committed a type II error, in which case more data would likely be needed to detect a potential difference. Alternatively, H~0~ might be true, i.e., there is no population difference to detect. If the p-value is below the alpha threshold, the result is instead interpreted as there being a significant difference between groups, inferring that there is an actual difference at the population level. Although many PET-studies likely suffer from insufficient power because of limited sample sizes, it is also important to note that a difference between groups may often be detectable before reaching the sample size that was determined *a priori*. Ideally, the researcher should include just the number of individuals needed to reach a correct conclusion, no more, no less. When performing more scans than needed, the PET researchers are wasting money and exposing people unnecessarily to radioactivity.

@@

If a researcher continues to collect data indefinitely, the default BF t-test will eventually converge to supporting the hypothesis most compatible with the population effect, i.e., it will show evidence for the null-hypothesis if in fact there is no mean difference, or for the alternative-hypothesis if there is an underlying true difference in the population [@Rouder2009]. However, such continuous collection of data is often unrealistic in a clinical PET study. Apart from the high cost, a PET examination also entails injecting subjects with radioactivity. Hence, a maximum sample size usually has to be decided on *a priori* and approved by an ethical review board and/or a radiation safety committee.

@@

Figure \ref{fig_k_at_5fpr}A and 3B show the estimated false positive rate when applying sequential BF testing in the case in which there is no difference between the two groups. Three different stopping thresholds have been used (BF > 3, 4 and 6, respectively). When the maximum allowed sample size increases, the false positive rate goes up. This is because the longer the sequential testing can go on, the more decisions are being made. Some of these decisions will be wrong, meaning that the researcher will claim support for an effect, when in fact there is no effect in the population. Eventually, when N becomes high enough, the rate of false positives will reach an asymptote (see Supplementary Figure S1). But with a maximum sample size above 30 subjects/group, it is not possible to keep false positives below the 5% using a decision threshold of 4. Instead, if the researchers wish to control the false positives at the 5% level, one option is to increase the decision threshold. Figure \ref{fig_k_at_5fpr}C and 3D show the estimated decision threshold that would be needed to keep the false positive rate at 5% when using sequential testing with the default BF t-test, while still starting at 12 subjects/group and checking the BF after the collection of each patient-control pair.

@@

##Objective 2 - stopping for both H~1~ and H~0~
The second aim of this tutorial is to assess whether sequential BF testing can be used to stop a clinical PET study early, both when there is an effect in the population, but also when the population effect is zero. After the collection of each data point, we will compare both the BF, and its reciprocal, to an *a priori* set decision threshold. If either the BF or its reciprocal passes the threshold, the study will be stopped, and we will declare support in favour of the alternative or null hypothesis, respectively.

###Simulation set-up 
We used the same simulation set-up as above, where a grid of maximum sample sizes and effect sizes ranging from 0 to 1.5 was used to evaluate sequential BF testing in a cross-sectional and paired design respectively. We used the same settings for the default BF t-test as above with one exception: the alternative hypothesis was still a Cauchy distribution (centered at zero with r=0.707) but now specified as being one-sided, instead of two-sided. This means that we anticipate that the effect will go in one direction (e.g., patients will have a higher BP~ND~ than controls), making the test into a one-sided default BF t-test. The reason for considering only a one-sided test in this scenario is that it is not possible to stop for H~0~ when using a two-sided test at smaller sample sizes when using reasonable decision thresholds (BF < $\nicefrac{1}{4}$, see Supplementary Figure S11).

@@

\begin{tcolorbox}[title=\textbf{Practical recommendations 2},title filled]
{In this simulation set-up we show that it is possible to stop a PET study early either when there is a true underlying effect in the population, or when the effect is zero. For a patient-control comparison, we recommend that sequential BF testing not start until at least 12 subjects per group have been collected, setting the BF threshold to a minimum of 3, and using a one-sided Cauchy distribution (centered around zero with a \textit{r} = 0.707) as the alternative hypothesis. Using a one-sided test means that the researcher must make an *a priori* prediction of which direction the effect will have, and not change this prediction after the study is started, similar to performing a classical one-sided NHST t-test.}
\end{tcolorbox}

@@

#### Evidence in favor of H~0~
One advantage of sequential BF testing is that a PET study can also be stopped early when there is no effect in the population, i.e., when H~0~ is true. If the researcher wants to use a sequential BF to assess support for both H~1~ and H~0~, we recommend using a one-tailed default BF t-test. This entails that the researcher must make an *a priori* prediction on the direction of the effect (e.g., patients have higher binding than control subjects, or vice versa). Without using one-sided default BF t-test, it will be difficult (or even impossible) to stop a study early in support for H~0~, given the sample sizes used in typical PET studies (N<100 subjects/group) while still using a threshold that is high enough to keep the rate of false evidence at reasonable levels (see Supplementary Figure S12).

@@

1. Increase the *a priori* set decision threshold for stopping (see Figure \ref{fig_k_at_5fpr}).
2. Recruit and scan a larger pool of subjects (>12/group) before initiating the sequential BF testing. 
3. Check BF less frequently, e.g., only after the collection of each second or third (etc.) comparison pair, instead of after each single one.

@@

Figure \ref{fig_powerPhase2} presents the true and false evidence rates the researchers can expect when using a one-sided BF t-test and when they are correct in their prediction of the direction of an effect. If there is a sizable patient-control difference in the population, but it goes in the opposite direction to the predicted one, then using this approach would lead studies to be stopped early due to the null-hypothesis being supported almost 100% of the time. Of course, the null is not true in this case, but it is closer to the true population effect than the *a priori* defined alternative hypothesis. Hence, it will often gain the most support from the data.

@@

#### Inconclusive findings 
For any given PET-study where sequential BF testing is applied there will be a chance of reaching Nmax without crossing the pre-set decision threshold. In such cases, where the researcher ends up with an “inconclusive” finding, the BF is still interpretable. The suggested evidence thresholds in Table 1 can be used to report the support in data in favor of the two competing hypotheses, even though BF never formally reached the *a priori* set stopping threshold. For example, if the BF is above 3 (or below $\nicefrac{1}{3}$), this can still be reported as “moderate” support in data for one hypothesis over the other. However, in doing so the researcher cannot any longer say that they are controlling the rate of false evidence at a prespecified level.

@@

If a PET researcher wants to perform a study using sequential BF testing, we recommend that they follow the steps outlined in a flow-chart found in supplementary information (Supplementary Figure S13). In order to perform the default BF t-test, the freely available *BayesFactor* package [@Morey2018] in R or point-and-click software JASP [@Quintana2018; @Love2019; @VanDoorn2019] can be used.
